\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[retainorgcmds]{IEEEtrantools}

\usepackage{marginnote}
\usepackage{endnotes}

\usepackage{fancyhdr}

%Listings stuff
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{color}

\definecolor{gray}{rgb}{0.5,0.5,0.5}
\lstset{
basicstyle={\small\ttfamily},
tabsize=3,
numbers=left,
numbersep=5pt,
numberstyle=\tiny\color{gray},
stepnumber=2,
breaklines=true
}

%Properly formatted differential 'd'
\newcommand{\ud}{\, \mathrm{d}}

%Format stuff
\pagestyle{fancy}
\headheight 35pt

%Header info
\chead{\Large \textbf{Sorting Complexity}}
\lhead{}
\rhead{}

\begin{document}
\section{Introduction}

	\begin{description}
		\item[Computational complexity.] Framework to study efficiency of algorithms for solving a particular problem.
		\item[Model of computation.] Allowable operations.
		\item[Cost model.] Operation count(s).
		\item[Upper bound.] Cost guarantee provided by some algorithm.
		\item[Lower bound.] Proven limit on cost guarantee of all algorithms.
		\item[Optimal algorithm.] Algorithm with best possible cost guarantee.
	\end{description}
	
	Sorting:
	\begin{itemize}
		\item Model of computation: decision tree (can access information only through compares)
		\item Cost model: number of comparisons
		\item Upper bound: $N\log N$ from mergesort
		\item Lower bound: $N\log N$
	\end{itemize}
	
	\subsection{Proving Lower Bound}
		
		\textbf{Proposition:} any compare-based sorting algorithm use at least $\log (N!) \approx N\log N$ compares in the worst case.
		\begin{itemize}
			\item Assume array consists of $N$ distinct values $a_1$ through $a_n$.
			\item Worst case dictated by height $h$ of decision tree.
			\item Binary tree of height $h$ has at most $2^h$ leaves.
			\item $N!$ different orderings $\rightarrow$ at least $N!$ leaves.
		\end{itemize}
		
		\begin{IEEEeqnarray}{rCl}
				2^h & \geq & \# leaves \geq N! \\
				h & \geq&  \log (N!) \approx N\log N
			\end{IEEEeqnarray}
		
%		\def\enotesize{\normalsize}
%		\theendnotes
\end{document}